Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-10 survey/survey.tex
major reorganization

%%%%%%%%%
\paragraph*{Programs as Reversible Deformations.}

To better understand the essence of ``conservation of information'' in
the context of computing, we first look for analogous ideas in
physics, but this time at the macro scale. Viewing information as a
physical object, what does it mean to transform an object in such a
way that we do not lose its fundamental character?

@@

\amr{***HERE***; operational semantics; It feels like I should be
  writing about how the combinators (both for spaces and for
  'permutations') work at a lower level. Did we ever write that up
  somewhere?  We have lots of Agda code which 'does it', nicely, but
  somehow I can't remember an English prose version of it. I am not
  sure how long we want to make it.  For example, do we talk about the
  'matrix' representation of combinators-for-permutations-on-[n], and
  how these make things like 0+a=a and associativity obvious by
  construction?
  Let’s have a geometric representation of spaces:
- A + B means putting A next to B
- A * B means putting A and B on two perpendicular axes and filling the square (volume) between them
All the combinators would indeed have a natural action and if we have
a few nice pictures it would be great. What do you think? }
\jc{You are correct, I don't think we have any English prose which
explains the operations of our combinators. The `matrix' representation
makes sense for the untyped version, but is not necessarily adequate for
the typed version, as it becomes quite an indirect way to view it, since you
first have to assign numbers to each 'path' through a type as a proxy.
Though if we want to have path-indexed instead of Fin-indexed matrices,
I would be fine. So we could indeed still draw those pictures.  However,
this would lead associativity of * to be an equation on two 3D matrices...}

@@

\jc{We should decide what you level convention is. Do we say
that $\iso$ is level 0, so that $\Leftrightarrow$ is level 1?
We need to be consistent about it.}

\jc{wavefront}

Operational Semantics:

@@

\begin{verbatim}
examples:
  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
  conditionals
  toffoli
  fredkin
\end{verbatim}

Implement Fredkin gate and mention the following:

@@

@@

\begin{figure}[t]
Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
\\
  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
\\
  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
\\
  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
\\
  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
\\
  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
\\
  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
\end{array}\]
\caption{\label{figj}Signatures of level-1 $\Pi$-combinators: associativity}
\end{figure}

\begin{figure}[t]
Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {((a \oplus b) \otimes c) \odot \dist \Leftrightarrow \dist \odot ((a \otimes c) \oplus (b \otimes c))}
\\
  {(a \otimes (b \oplus c)) \odot \distl \Leftrightarrow \distl \odot ((a \otimes b) \oplus (a \otimes c))}
\\
  {((a \otimes c) \oplus (b \otimes c)) \odot \factor \Leftrightarrow \factor \odot ((a \oplus b) \otimes c)}
\\
  {((a \otimes b) \oplus (a \otimes c)) \odot \factorl \Leftrightarrow \factorl \odot (a \otimes (b \oplus c))}
\end{array}\]
\caption{\label{figi}Signatures of level-1 $\Pi$-combinators: distributivity and factoring}
\end{figure}

\begin{figure}[t]
Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\idc \odot \, c \Leftrightarrow c}
\quad
  {c \, \odot \idc \, \Leftrightarrow c}
\quad
  {c\,\, \odot\,! c \Leftrightarrow \idc}
\quad
  {! c \odot c \Leftrightarrow \idc}
\\
  {c \Leftrightarrow c}
\quad
\Rule{}
  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
  {c_1 \Leftrightarrow c_3}
  {}
\quad
\Rule{}
  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
  {}
\end{array}\]
\caption{\label{figh}Signatures of level-1 $\Pi$-combinators: identity and composition}
\end{figure}

\begin{figure}[t]
Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\identlp \odot c \Leftrightarrow (c_0 \oplus c) \odot \identlp}
\qquad
  {\identrp \odot (c_0 \oplus c) \Leftrightarrow c \odot \identrp}
\\
  {\identlsp \odot c \Leftrightarrow (c \oplus c_0) \odot \identlsp}
\qquad
  {\identrsp \odot (c \oplus c_0) \Leftrightarrow c \odot \identrsp}
\\
  {\identlt \odot c \Leftrightarrow (c_1 \otimes c) \odot \identlt}
\qquad
  {\identrt \odot (c_1 \otimes c) \Leftrightarrow c \odot \identrp}
\\
  {\identlst \odot c \Leftrightarrow (c \otimes c_1) \odot \identlst}
\qquad
  {\identrst \odot (c \otimes c_1) \Leftrightarrow c \odot \identrst}
\\
  {\identlt \Leftrightarrow \distl \odot (\identlt \oplus \identlt)}
\\
\identlp \Leftrightarrow \swapp \odot \identrsp
\qquad
\identlt \Leftrightarrow \swapt \odot \identrst
\end{array}\]
\caption{\label{figg}Signatures of level-1 $\Pi$-combinators: unit}
\end{figure}

\begin{figure}[t]
Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\swapp \odot (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \odot \swapp}
\quad
  {\swapt \odot (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \odot \swapt}
\\
  {(\assocrp \odot \swapp) \odot \assocrp \Leftrightarrow ((\swapp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \swapp)}
\\
  {(\assoclp \odot \swapp) \odot \assoclp \Leftrightarrow ((\idc \oplus \swapp) \odot \assoclp) \odot (\swapp \oplus \idc)}
\\
  {(\assocrt \odot \swapt) \odot \assocrt \Leftrightarrow ((\swapt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \swapt)}
\\
  {(\assoclt \odot \swapt) \odot \assoclt \Leftrightarrow ((\idc \otimes \swapt) \odot \assoclt) \odot (\swapt \otimes \idc)}
\end{array}\]
\caption{\label{figf}Signatures of level-1 $\Pi$-combinators: commutativity and associativity}
\end{figure}

\begin{figure}[t]
Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
\[\def\arraystretch{1.3}
\begin{array}{c}
\Rule{}
  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
  {}
\qquad
\Rule{}
  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
  {}
\\
  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
\qquad
  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
\\
  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
\\
  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
\end{array}\]
\caption{\label{fige}Signatures of level-1 $\Pi$-combinators: functors}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \odot (\idc \oplus \, \identlp)}
\\
  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \odot (\idc \otimes \, \identlt)}
\end{array}\]
\caption{\label{figd}Signatures of level-1 $\Pi$-combinators: unit and associativity}
\end{figure}

\begin{figure}[t]
Let $c : t_1 \leftrightarrow t_2$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {(c \otimes \idc) \odot \absorbl \Leftrightarrow \absorbl \odot \idc}
\quad
  {(\idc \, \otimes c) \odot \absorbr \Leftrightarrow \absorbr \odot \idc}
\\
  {\idc \odot \, \factorzl \Leftrightarrow \factorzl \odot (\idc \otimes c)}
\quad
  {\idc \odot \, \factorzr \Leftrightarrow \factorzr \odot (c \otimes \idc)}
\\
  {\absorbr \Leftrightarrow \absorbl}
\\
  {\absorbr \Leftrightarrow (\distl \odot (\absorbr \oplus \absorbr)) \odot \identlp}
\\
  {\identlst \Leftrightarrow \absorbr}
\qquad
  {\absorbl \Leftrightarrow \swapt \odot \absorbr}
\\
  {\absorbr \Leftrightarrow (\assoclt \odot (\absorbr \otimes \idc)) \odot \absorbr}
\\
  {(\idc \otimes \absorbr) \odot \absorbl \Leftrightarrow (\assoclt \odot (\absorbl \otimes \idc)) \odot \absorbr}
\\
  {\idc \otimes \, \identlp \Leftrightarrow (\distl \odot (\absorbl \oplus \idc)) \odot \identlp}
\end{array}\]
\caption{\label{figc}Signatures of level-1 $\Pi$-combinators: zero}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{c}
  {((\assoclp \otimes \idc) \odot \dist) \odot (\dist \oplus \idc) \Leftrightarrow (\dist \odot (\idc \oplus \dist)) \odot \assoclp}
\\
  {\assoclt \odot \distl \Leftrightarrow ((\idc \otimes \distl) \odot \distl) \odot (\assoclt \oplus \assoclt)}
\end{array}\]
\vspace{ -0.5em}
\[\def\arraystretch{1.3}
\begin{array}{rcl}
  (\distl \odot (\dist \oplus \dist)) \odot \assoclp &\Leftrightarrow&
   \dist \odot (\distl \oplus \distl) \odot \assoclp ~\odot \\
&& (\assocrp \oplus \idc) ~\odot \\
&& ((\idc \oplus \swapp) \oplus \idc) ~\odot \\
&&      (\assoclp \oplus \idc)
\end{array}\]
\caption{\label{figb}Signatures of level-1 $\Pi$-combinators: associativity and distributivity}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{rcl}
  (\idc \otimes \swapp) \odot \distl &\Leftrightarrow& \distl \odot \swapp
\\
  \dist \odot (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \odot \distl
\end{array}\]
\caption{\label{figa}Signatures of level-1 $\Pi$-combinators: commutativity and distributivity}
\end{figure}

@@

@@

@@

\amr{*** another wavefront ***}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data IV: Circle ???}

Would be interesting

@@

Must mention infinite sets since we promised to do so in Sec 2

Discuss recursion, or add a section of infinite sets???

quest for other principles; other data models; graphs; HoTT

Data IV: Recursive Types

Add recursion to structured finite types, trace, feedback, loop...

@@

meta-circular interpreter

@@

Data V: Exclusive Disjunctions

@@

\jc{The reason I commented this out is that it is under-justified. The
reader will simply not understand what these next few lines are really
saying.}
Indeed one should take the physical principles underlying quantum mechanics,
the most successful physical theory known to us and adapt computation to
``learn'' from these principles. To illustrate the depth of our crisis, Scott
Aaronson, Umesh Vazirani, and others have proposed the following puzzle.

@@

%% Below are pieces of text that are not immediately useful above, but
%% from which we might still be able to pilfer usefully.

%%%%%%%%%
\subsection{Technical Overview}

The technical development derives from ideas which are implicit in the
works of \cite{toffoli:1980, Zuliani:2001:LR,
  malacaria2007assessing,ClarkHM07,Ghica:2007:GSS:1190216.1190269}.
(1) Functions whose output entropy is equal to their input entropy are
information preserving functions.  (2) A function
$f : b_1 \rightarrow b_2$ is \emph{logically reversible} if there
exists an inverse function $f^{\dagger}$ such that for any input
$v_1:b_1$ if there exists $v_2:b_2$ such that $f(v_1)=v_2$ \emph{iff}
$f^{\dagger}(v_2)=v_1$.  (3) Logically reversible functions are
information preserving.

To summarize:
\begin{enumerate}
\item Based on isomorphisms of sum and product
  types~\cite{Fiore:2004}, we develop a strong normalizing model of
  computing, which we call $\Pi$, that is logically reversible and
  information preserving. Every computation expressible in $\Pi$
  is an isomorphism.

\item We extend $\Pi$ with isorecursive type and trace operators from
  category theory to obtain $\Pi$. Every computation expressible in
  $\Pi$ is a partial isomorphism --- i.e. the system admits
  non-termination. $\Pi$ is Turing complete while being information
  preserving and logically reversible.

\item We develop the categorical semantics of these
  models~\cite{rc2011}. \amr{Mention n-categories.} We also develop
  a graphical notation for programming in these models that
  reminiscent of Penrose diagrams for
  categories~\cite{selinger-graphical}, Geometry of Interaction (GoI)
  machines and Proof
  Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
  notation, computation is modeled by the flow of particles in a
  circuit.

\item Since these new models are substantially different from \lcal\
  or familiar high-level languages, a point of concern is how one can
  effectively develop programs in them.  We address this by presenting
  a straightforward technique for deriving $\Pi$ programs by
  systematically translating logically reversible small-step abstract
  machines~\cite{isoint}. Complex programs can be devised in this way;
  we demonstrate the derivation of a meta-circular interpreter for
  $\Pi$.

\item We develop an arrow meta-language to encapsulate information
  effects. This metalanguage serves to encapsulate effects, in much
  the same way that traditional arrows or monads serve to encapsulate
  effects over the \lcal.

\item We show how a conventional irreversible model can be expressed
  in our model. We compile the first-order fragment of typed \lcal\
  extended with sums, products and loops to $\Pi$. This compilation is
  interesting for two reasons: (1) it exposes the implicit information
  effects of \lcal\ in the compilation to $\Pi$ and (2) the
  compilation of $\Pi$ to $\Pi$ shows that information effects can be
  erased by treating them as interactions with an explicit information
  environment.

\item We develop the notion of `duality' in $\Pi$ which gives us not
  one duality (like in linear logic or Filinski's symmetric
  \lcal~\cite{Filinski:89}), but two notions of duality -- an additive
  duality and a multiplicative duality. This gives a crisp semantics
  for negative and fractional types in the context of $\Pi$.

\end{enumerate}

Several intriguing areas of exploration remain and the book concludes
with discussion of some of these, along with other possible
applications.  Of these, the field of reversible and Quantum computing
is closely related and we survey several connections in depth.  The
exact connection between $\Pi$ and linear logic~\cite{Girard87tcs} is
not fully understood though it is clear that both systems capture
related but different notions of resource usage. A closely related
area is the connection with duality of computation \cite{Filinski:89,
  DBLP:conf/icfp/CurienH00, 10.1109/LICS.2010.23,Wadler:2003} ---
i.e., the quest for a unifying framework between `computations',
values, continuations and their logical counterparts. The
investigation of duality gives rise to negative and fractional types
which have an intuitive semantics and are reminiscent of negative
information flow, superposition and entanglement from Quantum
Physics~\cite{piee}.

%%%%%%%%%
\paragraph*{Conservation of Information.}

@@

% verbatim for now, will go to something decent later
\begin{verbatim}
check-password (guess) =
  if guess == "10"
  then True
  else False
\end{verbatim}

@@

%%%%%%%%%
\paragraph*{Information.}
We can now give a first precise definition of information preservation.
Let
`$b$' be a (not necessarily finite) type whose values are labeled
$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
as $-\sum p_i \log{p_i}$~\cite{shannon1948}.  Consider a function
$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
indicates the probability of the output of the function to have value
${b_2}^j$. We say a function is \emph{information-preserving} if its
output entropy is equal to the entropy of its input. In later sections,
we will refine this view.

\amr{
Talk about conservation of information in biology, computer science,
evolutionary computing, and search. (See
\url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.
}

In our current age, both Landauer's Principle and recent investigations
of computational models well-suited to quantum mechanics, push us to
revisit our current models of computation. Information, on top of
having a physical manifestation, appears to be \emph{conserved},
along with other quantities like energy, mass and momentum.

But what does it mean for a computation to preserve information?
What is the missing principle that will guide the creation of a new
model of computation that ``conserve information''?
It is nature to look again at physics for inspiration --
in our current understanding of physics all of the primitive rules
are reversible. Conventional computation is not. Can
we embrace this simple principle as the building block of a model of
computation? What would computation look like when viewed from this
vantage point? Can non-trivial computation be performed in a
reversible manner? And, if so, does it have applications?

Traditional models of computation are all unityped, generally
operating on unstructured sequences of symbols. But recent
investigations, in the intersection between mathematics and
computer science (particularly in Homotopy Type Theory~\cite{hottbook})
reveal that not only can mathematics be typed, but that doing so
reveals further structure -- in both mathematics and in computation.

And that brings us to the principal objective of this survey: explore the
domain of \emph{typed reversible computation}.

\amr{In the quantum case, this is clear. Say something about topological
quantum computing.}

\textbf{punch line: allowable operations that computers can perform
  while embracing the laws of physics are operations that preserve
  everthing in abstract space of ``data''; doughnut into cup kind of
  transformations.}

>>>>

%%%%%%%%%
\paragraph*{Programs as Reversible Deformations.}
To better understand the essence of ``conservation of information'' in
the context of computing, we first look for analogous ideas in
physics, but this time at the macro scale. Viewing information as a
physical object, what does it mean to transform an object in such a
way that we do not lose its fundamental character?

@@

@@

@@

%%%%%%%%%
\subsection{Operational Semantics}

%%%%%%%%%
\subsection{Graphical Language}

a graphical notation for programming in these models that reminiscent
of Penrose diagrams for categories~\cite{selinger-graphical}, Geometry
of Interaction (GoI) machines and Proof
Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
notation, computation is modeled by the flow of particles in a
circuit.

\amr{
  Let’s have a geometric representation of spaces:
- A + B means putting A next to B
- A * B means putting A and B on two perpendicular axes and filling
the square (volume) between them
All the combinators would indeed have a natural action and if we have
a few nice pictures it would be great. What do you think? }

Surprising expressiveness of Pi:

@@

\begin{verbatim}
examples:
  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
  conditionals
  toffoli
  fredkin
  revisit adder example from Sec 2 
  perhaps a larger example as well (from RC 18 paper?)
\end{verbatim}

@@

\begin{figure}[t]
Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
\\
  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
\\
  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
\\
  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
\\
  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
\\
  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
\\
  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
\end{array}\]
\caption{\label{figj}Signatures of level-2 $\Pi$-combinators: associativity}
\end{figure}

\begin{figure}[t]
Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {((a \oplus b) \otimes c) \odot \dist \Leftrightarrow \dist \odot ((a \otimes c) \oplus (b \otimes c))}
\\
  {(a \otimes (b \oplus c)) \odot \distl \Leftrightarrow \distl \odot ((a \otimes b) \oplus (a \otimes c))}
\\
  {((a \otimes c) \oplus (b \otimes c)) \odot \factor \Leftrightarrow \factor \odot ((a \oplus b) \otimes c)}
\\
  {((a \otimes b) \oplus (a \otimes c)) \odot \factorl \Leftrightarrow \factorl \odot (a \otimes (b \oplus c))}
\end{array}\]
\caption{\label{figi}Signatures of level-2 $\Pi$-combinators: distributivity and factoring}
\end{figure}

\begin{figure}[t]
Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\idc \odot \, c \Leftrightarrow c}
\quad
  {c \, \odot \idc \, \Leftrightarrow c}
\quad
  {c\,\, \odot\,! c \Leftrightarrow \idc}
\quad
  {! c \odot c \Leftrightarrow \idc}
\\
  {c \Leftrightarrow c}
\quad
\Rule{}
  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
  {c_1 \Leftrightarrow c_3}
  {}
\quad
\Rule{}
  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
  {}
\end{array}\]
\caption{\label{figh}Signatures of level-2 $\Pi$-combinators: identity and composition}
\end{figure}

\begin{figure}[t]
Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\identlp \odot c \Leftrightarrow (c_0 \oplus c) \odot \identlp}
\qquad
  {\identrp \odot (c_0 \oplus c) \Leftrightarrow c \odot \identrp}
\\
  {\identlsp \odot c \Leftrightarrow (c \oplus c_0) \odot \identlsp}
\qquad
  {\identrsp \odot (c \oplus c_0) \Leftrightarrow c \odot \identrsp}
\\
  {\identlt \odot c \Leftrightarrow (c_1 \otimes c) \odot \identlt}
\qquad
  {\identrt \odot (c_1 \otimes c) \Leftrightarrow c \odot \identrp}
\\
  {\identlst \odot c \Leftrightarrow (c \otimes c_1) \odot \identlst}
\qquad
  {\identrst \odot (c \otimes c_1) \Leftrightarrow c \odot \identrst}
\\
  {\identlt \Leftrightarrow \distl \odot (\identlt \oplus \identlt)}
\\
\identlp \Leftrightarrow \swapp \odot \identrsp
\qquad
\identlt \Leftrightarrow \swapt \odot \identrst
\end{array}\]
\caption{\label{figg}Signatures of level-2 $\Pi$-combinators: unit}
\end{figure}

\begin{figure}[t]
Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\swapp \odot (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \odot \swapp}
\quad
  {\swapt \odot (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \odot \swapt}
\\
  {(\assocrp \odot \swapp) \odot \assocrp \Leftrightarrow ((\swapp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \swapp)}
\\
  {(\assoclp \odot \swapp) \odot \assoclp \Leftrightarrow ((\idc \oplus \swapp) \odot \assoclp) \odot (\swapp \oplus \idc)}
\\
  {(\assocrt \odot \swapt) \odot \assocrt \Leftrightarrow ((\swapt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \swapt)}
\\
  {(\assoclt \odot \swapt) \odot \assoclt \Leftrightarrow ((\idc \otimes \swapt) \odot \assoclt) \odot (\swapt \otimes \idc)}
\end{array}\]
\caption{\label{figf}Signatures of level-2 $\Pi$-combinators: commutativity and associativity}
\end{figure}

\begin{figure}[t]
Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
\[\def\arraystretch{1.3}
\begin{array}{c}
\Rule{}
  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
  {}
\qquad
\Rule{}
  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
  {}
\\
  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
\qquad
  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
\\
  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
\\
  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
\end{array}\]
\caption{\label{fige}Signatures of level-2 $\Pi$-combinators: functors}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \odot (\idc \oplus \, \identlp)}
\\
  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \odot (\idc \otimes \, \identlt)}
\end{array}\]
\caption{\label{figd}Signatures of level-2 $\Pi$-combinators: unit and associativity}
\end{figure}

\begin{figure}[t]
Let $c : t_1 \leftrightarrow t_2$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {(c \otimes \idc) \odot \absorbl \Leftrightarrow \absorbl \odot \idc}
\quad
  {(\idc \, \otimes c) \odot \absorbr \Leftrightarrow \absorbr \odot \idc}
\\
  {\idc \odot \, \factorzl \Leftrightarrow \factorzl \odot (\idc \otimes c)}
\quad
  {\idc \odot \, \factorzr \Leftrightarrow \factorzr \odot (c \otimes \idc)}
\\
  {\absorbr \Leftrightarrow \absorbl}
\\
  {\absorbr \Leftrightarrow (\distl \odot (\absorbr \oplus \absorbr)) \odot \identlp}
\\
  {\identlst \Leftrightarrow \absorbr}
\qquad
  {\absorbl \Leftrightarrow \swapt \odot \absorbr}
\\
  {\absorbr \Leftrightarrow (\assoclt \odot (\absorbr \otimes \idc)) \odot \absorbr}
\\
  {(\idc \otimes \absorbr) \odot \absorbl \Leftrightarrow (\assoclt \odot (\absorbl \otimes \idc)) \odot \absorbr}
\\
  {\idc \otimes \, \identlp \Leftrightarrow (\distl \odot (\absorbl \oplus \idc)) \odot \identlp}
\end{array}\]
\caption{\label{figc}Signatures of level-2 $\Pi$-combinators: zero}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{c}
  {((\assoclp \otimes \idc) \odot \dist) \odot (\dist \oplus \idc) \Leftrightarrow (\dist \odot (\idc \oplus \dist)) \odot \assoclp}
\\
  {\assoclt \odot \distl \Leftrightarrow ((\idc \otimes \distl) \odot \distl) \odot (\assoclt \oplus \assoclt)}
\end{array}\]
\vspace{ -0.5em}
\[\def\arraystretch{1.3}
\begin{array}{rcl}
  (\distl \odot (\dist \oplus \dist)) \odot \assoclp &\Leftrightarrow&
   \dist \odot (\distl \oplus \distl) \odot \assoclp ~\odot \\
&& (\assocrp \oplus \idc) ~\odot \\
&& ((\idc \oplus \swapp) \oplus \idc) ~\odot \\
&&      (\assoclp \oplus \idc)
\end{array}\]
\caption{\label{figb}Signatures of level-2 $\Pi$-combinators: associativity and distributivity}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{rcl}
  (\idc \otimes \swapp) \odot \distl &\Leftrightarrow& \distl \odot \swapp
\\
  \dist \odot (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \odot \distl
\end{array}\]
\caption{\label{figa}Signatures of level-2 $\Pi$-combinators: commutativity and distributivity}
\end{figure}

@@

\begin{verbatim}
Mention convention:

level 0: types
level 1: type equivalences (pi combinators)
level 2: equivalences between equivalences

(cite MFPS invited paper)
\end{verbatim}

@@

%%%%%%%%%
\subsection{A Model of Equivalences between Type Equivalences}

@@

%%%%%%%%%
\subsection{A Language of Equivalences between Type Equivalences}

@@

%%%%%%%%%
\subsection{Operational Semantics}

%%%%%%%%%
\subsection{Graphical Language}

@@

Possible extensions; quest for other principles; other data models; graphs; HoTT

%%%
\subsection{Infinite Sets}

Must mention infinite sets since we promised to do so in Sec 2;
Discuss recursion, trace, feedback, loop...

@@

meta-circular interpreter from RC paper: Since these new models are
substantially different from \lcal\ or familiar high-level languages,
a point of concern is how one can effectively develop programs in
them.  We address this by presenting a straightforward technique for
deriving $\Pi$ programs by systematically translating logically
reversible small-step abstract machines~\cite{isoint}. Complex
programs can be devised in this way; we demonstrate the derivation of
a meta-circular interpreter for $\Pi$.

@@

%%%
\subsection{Topological Spaces}

Data IV: Circle etc. Would be interesting

%%%
\subsection{Theseus}

%%%
\subsection{Quantum Computing}

Extension of Theseus in 2018 FoSSaCS paper.

Exclusive Disjunctions

@@

\jc{The reason I commented this out is that it is under-justified. The
reader will simply not understand what these next few lines are really
saying.}

Indeed one should take the physical principles underlying quantum
mechanics, the most successful physical theory known to us and adapt
computation to ``learn'' from these principles. To illustrate the
depth of our crisis, Scott Aaronson, Umesh Vazirani, and others have
proposed the following puzzle.

@@

%%%
\subsection{Information Flow}

@@

\begin{verbatim}
check-password (guess) =
  if guess == "10"
  then True
  else False
\end{verbatim}

@@

%%%
\subsection{Conservation of Information, Information Effects, and Duality}

Talk about conservation of information in biology, computer science,
evolutionary computing, and search. (See
\url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.

We develop an arrow meta-language to encapsulate information
effects. This metalanguage serves to encapsulate effects, in much the
same way that traditional arrows or monads serve to encapsulate
effects over the \lcal.

new paper \url{https://arxiv.org/pdf/1805.08605.pdf}

Mention connections to negative, fractional, and algebraic numbers

%%%
\subsection{Summary}

\begin{comment}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theseus: Types and Simple Isomorphisms}
\label{theseusI}
